A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-14 paper.tex
Move phylo network discussion to end
Fixes #219

\textcolor{red}{TODO: reduce the text below down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature. I think pulling apart the
idea of a ``reticulation'' vs what we call a recombinant
is essential here.}

@@

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry} instead of focussing on events,
we define the ARG data structure in terms of \emph{genomes} and the
patterns of transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

@@

\section*{Evaluation of sampling probabilities}\label{probabilities}

@@

The conditional sampling probability $p(G | D)$ of an ARG $G$ given an observed
realization of genetic diversity $D$ at its sampled leaves is a central concept in
model fitting and selection. Under the Bayesian paradigm, it defines the posterior
distribution $p(G | D) \propto p(D | G) p(G)$, where $p(D | G)$ is the likelihood
of the data $D$ given $G$, and $p(G)$ is a prior distribution for the underlying
ARG $G$. Under the frequentist paradigm, $p(G)$ can be interpreted as a regularizer,
and $p(G | D)$ as a regularized likelihood. In both cases, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Essentially all of the model-based inference methods surveyed below
rely on evaluation of a likelihood and a prior. Two contemporary examples of
likelihood-based methods are \cite{mahmoudi2022bayesian} and \cite{guo2022recombination}.

In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is based on
a Poisson process on the ancestral material on the edges of the ARG, which implies
that mutation processes at two sites are conditionally independent given the
underlying ancestral trees at those sites. A prototypical example is
\citet[Eq.\ (2)]{mahmoudi2022bayesian}. Hence, $p(D | G)$ can be evaluated
regardless of the precise data structure used to store $G$, provided that it encodes
at least all marginal ancestral trees along the sampled sequences.  Choosing the
right data structure can still have a dramatic effect on the efficiency with which
$p(D | G)$ can be evaluated \citep{mahmoudi2022bayesian}.

In contrast, evaluating $p(G)$ under the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
the number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time along the ARG
\citep[Eq.\ (3)]{mahmoudi2022bayesian}. These cannot be uniquely determined
from just the marginal trees at each site because branches at multiple trees can be
contained in the same ancestral lineage. Instead, a richer data structure akin to the
eARG or gARG is needed. Moreover, because eARG does not explicitly store the flow
of ancestral material along its edges---a necessary piece of information for
computing the number of effective recombination links at each time---evaluating
$p(G)$ based on an eARG essentially requires turning it into a gARG at runtime.

The number of extant lineages at a given time is not explicitly represented even
in a gARG, and needs to be calculated from the node and edge tables.
For example, any edges with the same child and parent nodes are contained within
the same lineage. Under the coalescent with recombination, the two edges
originating from a recombinant child are also contained within a single lineage
until the time of their parent nodes, at which point they split. Nodes \textsf{s}, \textsf{t}, and \textsf{u}
in panel (a) of Figure \ref{fig-ancestry-resolution} provide a concrete example
of the latter: the edges from \textsf{s} to \textsf{t} and \textsf{s} to \textsf{u} are distinct, but correspond to only
a single lineage in the time interval they span. We emphasize that these
conventions grouping edges into lineages are not part of the gARG data structure.
While they are convenient for analysis using the coalescent with recombination,
other models, such as discrete time Wright--Fisher models with finite populations,
may benefit from different ways to associate edges with lineages.

@@

An alternative approach is to treat the ARG as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods operate on representations of the ``little ARG'', and are extremely computationally
expensive, being applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters. State-of-the-art methods rely on cheaper, approximate models
\citep{didelot2010inference, heine2018bridging, hubisz2020mapping,hubisz2020inference, medina2020speeding}.
The most scalable method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to dozens of mammal-like
genomes \citep{hubisz2020inference}.

Recently, \citet{mahmoudi2022bayesian} have improved the scalability of exact MCMC for ARGs by
leveraging the tree sequence representation of marginal trees, suitably enriched to facilitate the
evaluation of sampling probabilities. While not competitive with \texttt{ARGWeaver}, their method extends
the feasible range of sequence lengths by at least an order of magnitude.

@@

>>>>

@@

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry} instead of focussing on events,
we define the ARG data structure in terms of \emph{genomes} and the
patterns of transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

@@

@@

@@

An alternative approach is to treat the ARG as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg, mahmoudi2022bayesian}.
These methods operate on representations of the ``little ARG'', and are extremely computationally
expensive, being applicable to at most hundreds of samples consisting of tens or hundreds of kilobases with
human-like parameters. State-of-the-art methods rely on cheaper, approximate models
\citep{didelot2010inference, heine2018bridging, hubisz2020mapping,hubisz2020inference, medina2020speeding}.
The most scalable method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to dozens of mammal-like
genomes \citep{hubisz2020inference}.
A central quantity in all of these sampling methods is the conditional sampling probability 
$p(G | D) \propto p(D | G) p(G)$ of an ARG $G$ given an observed realization
of genetic diversity $D$ at its sampled leaves, where $p(D | G)$ is the likelihood
of the data $D$ given $G$, and $p(G)$ is a Bayesian prior distribution or a 
frequentist regularizer for the ARG $G$.

The likelihood $p(D | G)$ is effectively always the distribution of
a Poisson process on the ancestral material along the edges of $G$ \citep[Eq.\ (2)]{mahmoudi2022bayesian}.
Hence it is typically straightforward to evaluate in principle, but choosing the
right data structure can still have a dramatic effect on the efficiency of evaluation
\citep{mahmoudi2022bayesian}. Especially in the Bayesian case, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Two contemporary examples are
\cite{mahmoudi2022bayesian, guo2022recombination}.
Evaluating the distribution of the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time \citep[Eq.\ (3)]{mahmoudi2022bayesian}.
These require a data structure from which shared edges on different trees are easy to identify,
and which encodes recombination events and times explicitly, e.g.\ using nodes.

@@

\textcolor{red}{Discussion below on phylogenetic networks: we need to
reduce the text below down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature. I think pulling apart the
idea of a ``reticulation'' vs what we call a recombinant
is essential here.}
